A comparison of design-based and model-based approaches for spatial data.

### 2021-06-11 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
minor updates from unlink pdf issue

Finite Population Block Kriging (FPBK) is an alternative to sampling-based methods [@verhoef2008spatial]. FPBK expands the geostatistical kriging framework to the finite population setting. Instead of basing inference off of a specific sampling design, we assume the data were generated by a spatial process with parameters that can be estimated using the framework of a model.

@verhoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. For a response variable $\mathbf{z}$ that can be measured on a finite number of $N$ sites, we want to predict some linear function of the response variable, $\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}$ is a vector of weights. For example, if we want to predict the total abundance across all sites, then we would use a vector of 1's for the weights.

@@

@@

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $\tau(\mathbf{b}'\mathbf{z})$ $\tau(\mathbf{b}'\mathbf{z})$ is vague} and its prediction variance can be computed. While details of the derivation are in [@verhoef2008spatial], we note here that the predictor and its variance are both moment-based. Neither require a particular distribution for $\mathbf{z}$.

We only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. However, k-nearest-neighbors [@fix1951discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, can also be used to obtain predictions for a mean or total from spatially correlated responses in a finite population setting.

@@

@@

* both good: correctly specified model with high correlation
* break model: non-gaussian errors
* break design: small area estimation

@@

* change n or sampling fraction

* model-based: how should sample be drawn? should locations be fixed?

@@

>>>>

Finite Population Block Kriging (FPBK) is an alternative to sampling-based methods [@verhoef2008spatial]. FPBK expands the geostatistical kriging framework to the finite population setting. Instead of basing inference off of a specific sampling design, we assume the data are generated by a spatial process with parameters that can be estimated using the framework of a model.

<!-- don't like this: the framework of a model.  -->

@verhoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. For a response variable $\mathbf{z}$ that can be measured on a finite number of $N$ sites, we want to predict some linear function of the response variable, $\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}$ is a vector of weights. For example, if we want to predict the population total across all sites, then we would use a vector of 1's for the weights.

<!-- how explicitly do we want to define \mathbf{z} -->

<!-- finite number of -->

<!-- confusing having two tau's: tau for total and a tau as a function -->

@@

<!-- do we want the general model or do we want a constant mean model with a note that this can be expanded to include covariates? -->

@@

<!-- need to add stationarity to this definition -->

<!-- need to define indicator variable definition -->

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $\tau(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in [@verhoef2008spatial], we note here that the predictor and its variance are both moment-based. Neither require a particular distribution for $\mathbf{z}$.

<!-- a little weird to talk about inference being based on distributional assumptions and then throw in this sentence about not assuming a distribution for z. -->

We note that we only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. However, k-nearest-neighbors [@fix1951discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, can also be used to obtain predictions for a mean or total from spatially correlated responses in a finite population setting.

@@

__Sample Simulation__

@@

* both good: correctly specified model with high correlation
* break model: highly non-normal errors with small sample size
* break design: small area estimation

__Simulation Discussion Questions__

* model-based: how should sample be drawn? should locations be fixed?
* change n or sampling fraction?

__Other Base Settings?__

@@

@@

<!-- spsurvey is talked about in GRTS. so either move that down here or move this to FPBK. -->

### 2021-06-11 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
minor updates from unlink pdf issue

  \title{A comparison of design-based and model-based approaches for
spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matthew Higham\corref{1}}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

There are two general approaches for using data to make statistical
inferences about a population: design-based approaches and model-based
approaches. When data cannot be obtained for all units in a population
(known as population units), data on a subset of the population units is
collected in a sample. In the design-based approach, inferences about
the underlying population are informed from a probabilistic process in
which population units are selected to be in the sample. Alternatively,
in the model-based approach, inferences are made from specific
assumptions made about the underlying process that generated the data.
Each paradigm has a deep historical context (Sterba, 2009) and its own
set of general advantages (Hansen et al., 1983).

@@

The model-based approach assumes the data are a random realization of a
process. Randomness is often incorporated through distributional
assumptions on the data-generating process. Instead of estimating fixed
but unknown parameters (as in the design-based approach), the goal of
model-based inference in the spatial context is often \emph{prediction}
of an unknown quantity. For example, suppose the realized mean of all
population units is the quantity of interest. Instead of
\emph{estimating} a fixed unknown mean, we are \emph{predicting} the
value of the mean, a random variable. We know that if we sampled all
population units, we would have an exact prediction for the mean of our
one realized process, without any uncertainty. But the true mean of the
spatial process that generated our realized data is still not known.
When predicting the realized mean, we typically are not interested in
the underlying process' true mean.

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals can be more
precise. Cressie (1993) and Schabenberger and Gotway (2017) provide
reviews of model-based approaches for spatial data.

@@

Sampling designs can incorporate spatial locations to obtain samples
that are spatially balanced with respect to the population (Stevens Jr
and Olsen, 2004). A sample is spatially with respect to the population
if the sampled population units are a miniature of the population units.
A sample is a miniature of the population if the distribution of the
sampled population units mirrors the density of all population units.
Spatial balance with respect to the population is is different than
spatial balance with respect to geography. A sample that is spatially
balanced with respect to geography is spread out in some type of
equidistant manner over geographical space and are not meant to be
miniatures of the population. When we refer to spatial balance
henceforth, we mean spatial balance with respect to the population.

Spatially balanced samples are useful because they tend to yield
estimates that have lower variance than estimates constructed from
sampling designs lacking spatial balance (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens Jr and Olsen, 2004; Wang et al., 2013). To quantify
spatial balance, Stevens Jr and Olsen (2004) proposed loss functions
based on Voroni polygons. The first spatiallly balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (Stevens Jr and Olsen, 2004). Since GRTS was
developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning (Robertson et al., 2018) algorithms. We
focus on the Generalized Random Tessellation Stratified (GRTS) algorithm
to select spatially balanced sampling because the algorithm has several
attractive properties detailed by Stevens Jr and Olsen (2004) and
Dumelle et al. (2021).

@@

After collecting a sample, the data are used to estimate population
parameters. The Horvitz-Thompson estimator (Horvitz and Thompson, 1952)
and its continuous analog (Cordy, 1993) yield unbiased estimates of
population means and totals. If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\) (denoted by \(\hat{\tau}_{ht}\)),
is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) and \(pi_i\) are the observed value and
inclusion probability of the \(i\)th population unit selected in the
sample. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are included in the sample, which
can be very difficult to calculate. Second, they ignore the spatial
locations of the population units. To address these drawbacks, Stevens
Jr and Olsen (2003) proposed a local neighborhood variance estimator.
The local neighborhood variance estimator does not rely on \(\pi_{ij}\),
and it incorporates spatial locations by assigning higher weights to
nearby observations. Stevens Jr and Olsen (2003) show this variance
estimators tends to reduce the variability associated with estimating
\(\tau\). This yields confidence intervals for \(\tau\) that are
narrower than confidence intervals constructed from variance estimators
ignoring spatial locations.

@@

We only use FPBK in this paper in order to focus more on comparing the
design-based and model-based approaches. However, k-nearest-neighbors
(Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013), random forest
(Breiman, 2001), Bayesian models (Chan-Golston et al., 2020), among
others, can also be used to obtain predictions for a mean or total from
spatially correlated responses in a finite population setting.

Finite Population Block Kriging (FPBK) is an alternative to
samipling-based methods (Ver Hoef, 2008). FPBK expands the
geostatistical kriging framework to the finite population setting.
Instead of basing inference off of a specific sampling design, we assume
the data were generated by a spatial process with parameters that can be
estimated using the framework of a model.

Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. For a response variable
\(\mathbf{z}\) that can be measured on a finite number of \(N\) sites,
we want to predict some linear function of the response variable,
\(\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}\) is a vector of weights. For example, if we want to
predict the total abundance across all sites, then we would use a vector
of 1's for the weights.

@@

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(\tau(\mathbf{b}'\mathbf{z})\)
\(\tau(\mathbf{b}'\mathbf{z})\) is vague\} and its prediction variance
can be computed. While details of the derivation are in (Ver Hoef,
2008), we note here that the predictor and its variance are both
moment-based. Neither require a particular distribution for
\(\mathbf{z}\).

@@

@@

\begin{itemize}
\item
  both good: correctly specified model with high correlation
\item
  break model: non-gaussian errors
\item
  break design: small area estimation
\item
  both good?: misspecified covariance model with high correlation
\item
  break both? non-gaussian areas with smaller sample size
\item
  change n or sampling fraction
\item
  model-based: how should sample be drawn? should locations be fixed?
\end{itemize}

>>>>

  \title{A comparison of design-based and model-based approaches for
spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matt Higham\corref{1}}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

There are two general approaches for using data to make statistical
inferences about a population: design-based approaches and model-based
approaches. When data cannot be obtained for all units in a population
(known as population units), data on a subset of the population units is
collected in a sample. In the design-based approach, inferences about
the underlying population are informed from a probabilistic process in
which population units are selected to be in the sample. Alternatively,
in the model-based approach, inferences are made from specific
assumptions about the underlying process that generated the data. Each
paradigm has a deep historical context (Sterba, 2009) and its own set of
general advantages (Hansen et al., 1983).

@@

The model-based approach assumes the data are a random realization of a
process. Randomness is often incorporated through distributional
assumptions on the data-generating process. Instead of estimating fixed
but unknown parameters (as in the design-based approach), the goal of
model-based inference in the spatial context is often \emph{prediction}
of an unknown quantity. For example, suppose the realized mean of all
population units is the quantity of interest. Instead of
\emph{estimating} a fixed unknown mean, we are \emph{predicting} the
value of the mean, a random variable. We know that if we sampled all
population units, we would have an exact prediction for the mean of our
one realized process, without any uncertainty. But the true mean of the
spatial process that generated our realized data is still not known.
When predicting the realized mean, we typically are not interested in
the underlying process's true mean.

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals to be more
precise. Cressie (1993) and Schabenberger and Gotway (2017) provide
reviews of model-based approaches for spatial data.

@@

Sampling designs can incorporate spatial locations to obtain samples
that are spatially balanced with respect to the population (Stevens Jr
and Olsen, 2004). A sample is spatially with respect to the population
if the sampled population units are a miniature of the population units.
A sample is a miniature of the population if the distribution of the
sampled population units mirrors the density of all population units.
Spatial balance with respect to the population is different than spatial
balance with respect to geography. A sample that is spatially balanced
with respect to geography is spread out in some type of equidistant
manner over geographical space and is not meant to be miniatures of the
population. When we refer to spatial balance henceforth, we mean spatial
balance with respect to the population.

Spatially balanced samples are useful because they tend to yield
estimates that have lower variance than estimates constructed from
sampling designs lacking spatial balance (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens Jr and Olsen, 2004; Wang et al., 2013). To quantify
spatial balance, Stevens Jr and Olsen (2004) proposed loss functions
based on Voroni polygons. The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (Stevens Jr and Olsen, 2004). Since GRTS was
developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning (Robertson et al., 2018) algorithms. We
focus on the Generalized Random Tessellation Stratified (GRTS) algorithm
to select spatially balanced sampling because the algorithm has several
attractive properties detailed by Stevens Jr and Olsen (2004) and
Dumelle et al. (2021).

@@

After collecting a sample, the data are used to estimate population
parameters. The Horvitz-Thompson estimator (Horvitz and Thompson, 1952)
and its continuous analog (Cordy, 1993) yield unbiased estimates of
population means and totals. For example, if \(\tau\) is a population
total, then the Horvitz-Thompson estimator of \(\tau\) (denoted by
\(\hat{\tau}_{ht}\)), is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) and \(\pi_i\) are the observed value and
inclusion probability of the \(i\)th population unit selected in the
sample. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are included in the sample, which
can be very difficult to calculate. Second, they ignore the spatial
locations of the population units. To address these drawbacks, Stevens
Jr and Olsen (2003) proposed a local neighborhood variance estimator.
The local neighborhood variance estimator does not rely on \(\pi_{ij}\),
and it incorporates spatial locations by assigning higher weights to
nearby observations. Stevens Jr and Olsen (2003) show this variance
estimators tends to reduce the variability associated with estimating
\(\tau\). This yields confidence intervals for \(\tau\) that are
narrower than confidence intervals constructed from variance estimators
ignoring spatial locations.

@@

Finite Population Block Kriging (FPBK) is an alternative to
sampling-based methods (Ver Hoef, 2008). FPBK expands the geostatistical
kriging framework to the finite population setting. Instead of basing
inference off of a specific sampling design, we assume the data are
generated by a spatial process with parameters that can be estimated
using the framework of a model.

Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. For a response variable
\(\mathbf{z}\) that can be measured on a finite number of \(N\) sites,
we want to predict some linear function of the response variable,
\(\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}\) is a vector of weights. For example, if we want to
predict the population total across all sites, then we would use a
vector of 1's for the weights.

@@

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(\tau(\mathbf{b}'\mathbf{z})\) and its prediction variance
can be computed. While details of the derivation are in (Ver Hoef,
2008), we note here that the predictor and its variance are both
moment-based. Neither require a particular distribution for
\(\mathbf{z}\).

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. However,
k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013),
random forest (Breiman, 2001), Bayesian models (Chan-Golston et al.,
2020), among others, can also be used to obtain predictions for a mean
or total from spatially correlated responses in a finite population
setting.

@@

\textbf{Sample Simulation}

@@

\begin{itemize}
\tightlist
\item
  both good: correctly specified model with high correlation
\item
  break model: highly non-normal errors with small sample size
\item
  break design: small area estimation
\end{itemize}

\textbf{Simulation Discussion Questions}

\begin{itemize}
\tightlist
\item
  model-based: how should sample be drawn? should locations be fixed?
\item
  change n or sampling fraction?
\end{itemize}

\textbf{Other Base Settings?}

\begin{itemize}
\tightlist
\item
  both good?: misspecified covariance model with high correlation
\item
  break both? non-gaussian areas with smaller sample size
\end{itemize}
